---
title: "ED 2.2 R interface"
author: "Alexey Shikloamnov"
output:
  rmarkdown::github_document
---

# Prepare inputs

For this tutorial, in your working directory, create a folder called `ed-input-data`.

Download meteorology inputs from the [GitHub releases page][releases].
For this tutorial, use the `NARR-ED2.tar.gz` file (North American Regional Reanalysis).
Download the file into `ed-input-data` and extract it.

```sh
cd ed-input-data
tar -xf NARR-ED2.tar.gz
cd ..
```

[releases]: https://github.com/FoRTExperiment/ed4forte/releases/tag/met-ed2

You should now have a directory `ed-input-data/NARR-ED2` containing a bunch of files like `1979JAN.h5` as well as a file called `ED_MET_DRIVER_HEADER`.

Open the `ed-input-data/NARR-ED2/ED_MET_DRIVER_HEADER` file in a text editor, and change the third line (`/Users/shik544/...`) so it matches the path of the directory
(e.g. `/path/to/current/directory/ed-input-data/NARR-ED2/`).
You can also do this with the following shell command:

```sh
# Use `gsed` instead of `sed` if on MacOS
sed -i "3s:.*:$PWD/ed-input-data/NARR-ED2/:" ed-input-data/NARR-ED2/ED_MET_DRIVER_DATA
```

The remaining inputs required for a basic ED2 simulation ship with the `ed4forte` package.
Install the package from your local clone (`devtools::install("/path/to/ed4forte")`) or from GitHub (`devtools::install_github("FoRTExperiment/ed4forte")`).

# Basic ED2 run

First, locate the ED2 executable and set the R option `ed4forte.ed2_exe` to the absolute path to this executable.

```{r set-option-show, eval = FALSE}
options(ed4forte.ed2_exe = "/path/to/ED/build/ed_2.2-dbg")
```

```{r set-option-hide, include = FALSE}
options(ed4forte.ed2_exe = "/Users/shik544/Projects/edmodel/ed2-my-develop/ED/build/ed_2.2-opt",
        width = 80)
```

Use the following R code to perform a simple ED2 simulation at UMBS for the year 2000 starting from bare ground:

```{r fix-paths, include = FALSE}
ed_input_dir <- here::here("unsynced-data", "ed-input-data")
stopifnot(file.exists(file.path(ed_input_dir, "NARR-ED2", "ED_MET_DRIVER_HEADER")))
```

```{r run-ed-hide, include = FALSE}
library(ed4forte)
outdir <- "test-ed-outputs"
if (length(list.files(outdir, "analysis-E")) == 12) {
  # Just sleep for a few seconds, so we have a process object to inspect
  p <- processx::process$new("sleep", "3")
} else {
  # Actually run ED
  p <- run_ed2(
    outdir,
    "2000-01-01",
    "2001-01-01",
    ED_MET_DRIVER_DB = file.path(ed_input_dir, "NARR-ED2", "ED_MET_DRIVER_HEADER")
  )
}
```

```{r run-ed-show, eval = FALSE}
library(ed4forte)

outdir <- "test-ed-outputs"
ed_input_dir <- "ed-input-data"

p <- run_ed2(
  outdir,
  "2000-01-01",
  "2001-01-01",
  ED_MET_DRIVER_DB = file.path(ed_input_dir, "NARR-ED2", "ED_MET_DRIVER_HEADER")
)
```

The first argument to `run_ed2` is the output directory (which will be created if it doesn't exist).
The next two arguments are the start and end date (-times), respectively.
The last argument indicates that the default ED2IN value of the `ED_MET_DRIVER_DB` tag should be modified to that value.
Any default values for ED2IN tags can be modified in this way.

Note that the code above will return immediately.
This is because it triggers ED2 to run in the background.
The return object (`p`) is a [`processx`](https://processx.r-lib.org/index.html) process, which can be examined.

```{r examine-status}
p$get_status()
p$is_alive()
```

For more details, see the [`processx::process` documentation](https://processx.r-lib.org/reference/process.html).
To wait for the run to complete, use the `wait()` method.

```{r wait-for-ed}
p$wait()
```

You can tell that the run finished by examining the contents of the output directory.

```{r list-output-files}
list.files(outdir)
```

You should see the 12 monthly output files, along with a copy of the ED2IN file used for the run and two log files corresponding to the input (`stdout`) and error (`stderr`) streams.

# Reading output

ED2 outputs are in HDF5 format, of which NetCDF is a special case.
Therefore, they can be read by NetCDF utilities, like the `ncdf4` package.

```{r ncdf4-outputs}
outfiles <- list.files(outdir, "analysis-E-", full.names = TRUE)
nc <- ncdf4::nc_open(outfiles[1])
ncdf4::ncvar_get(nc, "AGB_CO")
ncdf4::ncvar_get(nc, "MMEAN_GPP_PY")
```

Unfortunately, ED2's produces only one file per timestep, which makes reading in results from long simulations  pain.
`ed4forte` provides some utilities for reading a subset of ED2 file types and variables in bulk.

```{r read-outputs, warning = FALSE}
results <- read_monthly_dir(outdir)
results

scalar_results <- tidyr::unnest(results, df_scalar)
scalar_results

plot(MMEAN_GPP_PY ~ datetime, data = scalar_results, type = "o",
     main = "Total plot GPP")

pft_results <- tidyr::unnest(results, df_pft)
pft_results

plot(MMEAN_LAI_PY ~ datetime, data = pft_results, col = pft, pch = 19,
     main = "LAI by PFT")
```

# Configuring ED2 runs

ED2 runs are configured using two files:
`ED2IN` is a Fortran namelist file generally used for settings that affect the entire run (e.g. start and end dates; integration timestep; submodel configurations).
A reference `ED2IN` with documentation and default values (lightly tweaked for FoRTE) is in the [`inst/ED2IN`](../inst/ED2IN) file.
`config.xml` is an XML file that contains PFT-specific parameters, plus a few others.
A complete example featuring all of the ED2 default values can be found in the [`inst/ed2-default-parameters.xml`](../inst/ed2-default-parameters.xml) file.
Confusingly, some parameters can be set from both the `config.xml` and the `ED2IN` -- where both are used, parameters in the `config.xml` override the `ED2IN` values.

Both sets of parameters can be modified through the `run_ed2` function.
To modify `ED2IN` values, pass the name of the tag (without the leading `NL%`) directly as an argument to `run_ed2`.
For instance, in the below example, we modify the `ED_MET_DRIVER_HEADER` and `CROWN_MOD` tags.

```{r, eval = FALSE}
run_ed2(
  outdir, start_dt, end_dt,
  ED_MET_DRIVER_HEADER = "/path/to/ED_MET_DRIVER_HEADER",
  CROWN_MOD = 1
)
```

To see the default values of the `ED2IN` file, and for some documentation about what they mean, 

The parameter configuration XML file (typically called `config.xml`) can be configured via the `configxml` argument, which can be either a path to a complete `config.xml` file or (probably easier) a `list` of values.
For example:

```{r, eval = FALSE}
run_ed2(
  outdir, start_dt, end_dt,
  configxml = list(
    pft = list(num = 9, SLA = 35),
    pft = list(num = 10, Vm0 = 27.4),
    phenology = list(theta_crit = -1.25)
  ),
  ED_MET_DRIVER_HEADER = "/path/to/ED_MET_DRIVER_HEADER"
)
```

Note the format: A nested list of lists, with the name of each first-level list indicating the _type_ of parameter and each sub-list consisting of name-value pairs.
In the above example, we are setting the SLA of PFT number 9 (Temperate Early-successional Hardwood) to 35 and the Vcmax of PFT 10 (Temperate Mid-successional Hardwood) to 27.4, as well as setting the PFT-independent phenology submodel parameter `theta_crit` to -1.25.

Because the most common use of the `config.xml` is setting PFT parameters, which may be well-suited to a tabular data structure, you can also pass parameters as a `data.frame`
(but only if you're _only_ setting PFT parameters --- if you have a mix of PFT parameters and other values, you need to give the explicit nested list.
Also, for this to work, the same set of PFT parameters has to be set for all PFTs.):

```{r, eval = FALSE}
run_ed2(
  outdir, start_dt, end_dt,
  configxml = data.frame(
    num = c(9, 10),
    SLA = c(35.3, 38.2),
    Vm0 = c(27.4, 24.3)
  ),
  ED_MET_DRIVER_HEADER = "/path/to/ED_MET_DRIVER_HEADER"
)
```

If ED2 finds and reads a `config.xml` file, it will also write out a file called `history.xml` in the same directory that contains a complete list of all parameters and values used.
This is useful for figuring out what ED2's default parameters are, and confirming that parameters from `config.xml` have been read in correctly.
(The `history.xml` file from a test run is actually the basis for the `ed2-default-parameters.xml` file that ships with this package.
Also, for a tabular version of the default PFT parameters, see the [inst/ed2-default-pft-parameters.csv](../inst/ed2-default-pft-parameters.csv) file).

**NOTE**: A key difference between the `ED2IN` and `config.xml` files is how strictly they are parsed and how values unknown to ED2 are treated.
The `ED2IN` file must be complete and have no values that ED2 does not know about -- both missing entries and entries that ED2 doesn't know about will result in errors during run initialization.
On the other hand, unknown values in `config.xml` are _silently ignored_ -- i.e. ED2 will only read in parameters that it knows about.
This is why it's a good idea to check the `history.xml` file to make sure that your parameters from the `config.xml` are being correctly set.
